---
title: "R Explorations"
output: html_notebook
---

In this notebook, I am exploring certain R concepts to help with a better understanding of the language. These code bits are also here for reference. 

Index:
* Useful functions
  - download files
* ggplot histogram
* Latex fundamentals
* t-statistic

## Resources
* http://www.statmethods.net/index.html
* 

## Useful functions: 
#### Setup
```{r}
library(readr)
library(magrittr)
library(tidyr)
library(data.table)
library(ggplot2)
```
#### Download url data

```{r comment="download"}
fetch_data <- function(data_url, type="delim") {
  if (type=="delim") {
    my_data <- read_delim(data_url, " ", escape_double = FALSE, trim_ws = TRUE)    
  } else if (type == "table") {
    my_data <- read.table(data_url, sep=",", header=T)
  } else if (type == "csv") {
    my_data <- read.csv(data_url)
  } else {
    my_data = NULL
  }
  return(my_data)
}
```


#### Latex


#### Plotting
Some practice with creating and plotting dataframes in R

```{r}
df = data.frame(time = 1:10,
                a = cumsum(rnorm(10)),
                b = cumsum(rnorm(10)),
                c = cumsum(rnorm(10))
)
dfm <- melt(df, id.vars = 'time', variable.name = 'series', value.name = 'meas')

# Two ways of plotting:
ggplot(dfm, aes(time, meas)) + geom_line(aes(color=series))
ggplot(dfm, aes(time, meas)) + geom_line() + facet_grid(series ~ .)
```

#### t-test (from help)
```{r}
## Classical example: Student's sleep data
plot(extra ~ group, data = sleep)
## Traditional interface
with(sleep, t.test(extra[group == 1], extra[group == 2]))
## Formula interface
t.test(extra ~ group, data = sleep)
#View(sleep)
```